Thesis Title

### 2022-03-26 chapters/introduction.tex
finish acronym list

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. 
\ac{PCR} and reverse transcription-PCR (RT-PCR) are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of and demand for prompt information about viral loads, viral sequence data and potential antiviral resistance information \citep{cassedy_virus_2021}. 
Syndromic testing (see \secref{sssec:_intro_bacterial}) is now fully integrated into standard testing practises of many clinical laboratories \citep{dien_bard_panels_2020}. 
The limitations of these assays include the absence of detection of off-target pathogens, a lack of full susceptibility information, cost, and false positive results. 
\ac{qPCR} remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 5.png}
\caption{\textbf{The three revolutions in sequencing technology that have transformed the landscape of bacterial genome sequencing.} The first-generation, also known as Sanger sequencers, is represented by the ABI Capillary Sequencer (Applied Biosystems). During the sequencing reaction, at each nucleotide incorporation event, a fluorescently labelled \ac{ddNTP} is incorporated, terminating the elongation of the DNA molecule. The resulting electropherogram for sequencing reaction is below, and is read from left to right. The second-generation, also known as high-throughput sequencers, is represented by MiSeq, a 4-channel sequencer, and NextSeq, a 2-channel sequencer (Illumina), both sequencing by synthesis instruments. For both instruments, the loaded flowcell is sequenced in massive parallel reactions, with each nucleotide incorporation emitting a light signal that is captured and latter basecalled into a fastq file, with indication of the confidence of the call, presented below. In a 4-channel instrument each nucleotide has its own marker (A: yellow, T: green, C: red, G: blue) but in a 2-channel instrument only 2 markers exist (A: green plus red, T: green, C: red, G: no marker). These instruments allow the sequencing of both ends of the DNA fragment. Lastly, the third-generation, also known as long-read sequencers, is represented by the Pacific Bioscience BS sequencer and Oxford Nanopore MinION sequencer. In the first, immobilised polymerases in a SMRT Cell incorporating nucleotides with identifying fluorescent labels. In the latter, a nanopore embedded in a solid-state membrane causes a change in an ionic current across the membrane each time a nucleotide is pushed through the pore. This difference in potential is then used for basecalling. Adapted from \cite{hagemann_overview_2015, loman_twenty_2015,goodwin_coming_2016, wang_nanopore_2021, metzker_sequencing_2010, xu_recent_2020}}
\label{fig:figure5}
\end{figure*}

@@

The basis of Pacific Biosciences sequencers is known as single-molecule real-time sequencing (SMRT), which takes place in single-use SMRT Cells. 
These contain multiple immobilised polymerases, which, after binding to an adaptor sequence, begin replication incorporating nucleotides with identifying fluorescent labels. 
The sequence of fluorescence pulses is recorded into a movie which is then converted into a nucleotide sequence. 
After the polymerase completes replication of one DNA strand, it continues to sequence the opposite adapter and second strand. 
As a result, multiple passes of the same template can be generated depending on the lifetime of the polymerase \citep{hoang_long-reads-based_2022, loman_twenty_2015}. 
This technology has accuracy comparable with the Illumina systems but requires a higher initial investment cost, are much larger machines in comparison with the benchtop counterparts, and have much lower throughput and longer library preparation protocols \citep{hoang_long-reads-based_2022, wenger_accurate_2019}.

@@

Whole-genome sequencing (WGS) is becoming one of the most widely used applications of microbial genome sequencing. 
The major advantage of WGS is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:_intro_metagenomics}). 
In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:_intro_bacterial}) or viral (see \secref{sssec:_intro_viral}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures (see Section \ref{sec:_intro_bioinformatics}), but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, nature_reviews_genetics_genomic_2019, goodwin_coming_2016}.

@@

WGS has been used in the routine laboratory workflow when typing of pathogens by a method having the highest possible discriminatory power is required either through single nucleotide polymorphism (SNP) or core-genome/whole genome MLST (cg/wg MLST) analysis, for example during hospital outbreaks \citep{tagini_bacterial_2017}.

The implementation of WGS in routine diagnostics requires several adaptations in the laboratory workflow, from the ‘wet’ laboratory part (extraction, library preparation, sequencing), to the ‘dry’ bioinformatics part where genomic data is analysed and its results interpreted by specialised personnel \citep{rossen_practical_2018}.

@@

Most notably, WGS has become a common tool in surveillance and infection prevention, allowing for pathogen identification and tracking, establishing transmission routes and outbreak control \citep{lo_genomics_2020}. 
In bacterial infections, initiatives such as Pathogenwatch\footnote{\url{https://pathogen.watch/}} offers a web-based platform for \ac{AMR} analysis and phylogeny generation of \textit{Campylobacter}, \textit{Klebsiella}, \textit{Neisseria gonorrhoeae}, \textit{Staphylococcus aureus}, and \textit{Salmonella Typhi} \citep{afolayan_overcoming_2021}. 
The Center for Genomic Epidemiology website\footnote{\url{https://www.genomicepidemiology.org/}} offers services for phylogenetic tree building and \ac{AMR} prediction. 
Chewie Nomenclature Server\footnote{\url{https://chewbbaca.online/}} allows users to share genome-based gene-by-gene typing schemas and to maintain a common nomenclature, simplifying the comparison of results \citep{mamede_chewie_2021}. 
Enterobase\footnote{\url{https://enterobase.warwick.ac.uk/}} allows for the analysis and visualisation of genomic variation within enteric bacteria \citep{zhou_enterobase_2020}. 
Microreact\footnote{\url{https://microreact.org/}}, from the same developers as Pathogenwatch, combines clustering, geographical and temporal data into an interactive visualisation with trees, maps, timelines and tables for a multitude of microorganisms, both bacterial and viral \citep{argimon_microreact_nodate}. 
Particularly for viruses, GISAID\footnote{\url{https://www.gisaid.org/}}  promotes the rapid sharing of data from all influenza viruses and the coronavirus causing COVID-19, including the genetic sequences and related clinical and epidemiological data \citep{shu_gisaid_2017}. 
ViPR\footnote{\url{https://www.viprbrc.org/}} provides access to sequence records, gene and protein annotations, immune epitopes, 3D structures, host factor data, and other data types for over 14 viral families, including \textit{Coronaviridae}, from which \ac{SARS-CoV-2} belongs to, and \textit{Faviviridae}, the family of Dengue and Zika virus \citep{pickett_virus_2012}. 
INSaFLU\footnote{\url{https://insaflu.insa.pt/}} supplies public health laboratories and influenza researchers with a web-based suite for effective and timely influenza and \ac{SARS-CoV-2} laboratory surveillance, identifying the type and subtype/lineage, detection of putative mixed infections and intra-host minor variants \citep{borges_insaflu_2018}.

@@

Molecular barcoding approaches can be combined with second-generation high-throughput sequencing to achieve unprecedented depths of coverage in microbial community profiling, being defined as metataxonomics. 
For profiling bacterial species, the most popular approach is 16S \ac{rRNA} gene sequencing, an ~1500 base pair gene coding for a catalytic RNA that is part of the 30S ribosomal subunit. 
Traditionally, the variable regions of the 16S \ac{rRNA} gene (V-regions) are targeted, or ranges thereof (V1-V2, V1-V3, V3-V4, V4, V4-V5, V6-V8, and V7-V9), and are specific to bacterial genus (96\%) and for some, even species (87.5\%), \citep{srinivasan_use_2015, abellan-schneyder_primer_2021}. 
Moreover, dedicated 16S databases that include near full length sequences for a large number of strains and their taxonomic placements exist, such as RDP\footnote{\url{http://rdp.cme.msu.edu/}}, Greengenes\footnote{\url{https://greengenes.secondgenome.com/}}, silva\footnote{\url{https://www.arb-silva.de/}} and NCBI's 16S ribosomal RNA project\footnote{\url{https://www.ncbi.nlm.nih.gov/refseq/targetedloci/}} \citep{cole_ribosomal_2009, desantis_greengenes_2006, pruesse_silva_2007}. 
The sequence from an unknown strain can be compared against the sequences in these databases, after very closely related sequences are grouped into Operational Taxonomic Units (OTUs), and infer likely taxonomy, with the assumption that sequences of $>$95\% identity represent the same genus, whereas sequences of $>$97\% identity represent the same species \citep{schloss_introducing_2005}. 
Additionally, NCBI also provides the 23S ribosomal RNA project for Bacteria and Archaea metataxonomics.

Additionally, it must necessarily account for intragenomic variation between 16S gene copies. 
Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. 
Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 \ac{bp}) gene \citep{johnson_evaluation_2019}. 
The emergence of third generating sequencing technologies (see \secref{ssec:_intro_3rd_gen_seq}) allows for this limitation to be overcome but currently, only a fraction of the databases includes complete 16S \ac{rRNA} sequences.

@@

Shotgun metagenomics can offer relatively unbiased pathogen detection and characterisation. The capacity to detect all potential pathogens — bacteria, viruses, fungi and parasites — in a sample has great potential utility in the diagnosis of infectious disease \citep{chiu_clinical_2019}, potentially able to provide genotyping, antimicrobial resistance and virulence profiling in a single methodological step. This comes with the cost of producing massive amounts of information that require expert handling and processing, as well as capable computational infrastructures \citep{couto_critical_2018, rossen_practical_2018}.

Clinical applications of shotgun metagenomics derive its roots from the use of microarrays (see \secref{ssec:_intro_current_standards}), where it was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of second-generation sequencing technology and its high throughput of genomic data at a competitive price that made the sequencing of all genomic content, DNA and/or RNA) if a clinical sample a viable possibility for diagnostics (see \secref{ssec:_intro_2nd_gen_seq}) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of shotgun metagenomics was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since many other reports of successful application of shotgun metagenomics in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

In public health microbiology, shotgun metagenomics combined with transmission network analysis allowed the investigation and quick action on the food supply of the 2013 outbreak of Shiga toxin-producing \textit{Escherichia coli} (STEC) strain O104:H4 from faecal specimens obtained from patients \citep{loman_culture-independent_2013}. A similar approach was followed in the detection of \textit{Salmonella enterica} subsp. \textit{enterica} serovar Heidelberg from faecal samples in two though to be unrelated outbreaks in the United States of America, as well as the \textit{in situ} abundance and level of intrapopulation diversity of the pathogen, and the possibility of co-infections with \textit{Staphylococcus aureus}, overgrowth of commensal \textit{Escherichia coli}, and significant shifts in the gut microbiome during infection relative to reference healthy samples \citep{huang_metagenomics_2017}. More recently, shotgun metagenomic sequencing has evidenced alterations in the gut microbiota of a subset of COVID-19 patients that present the uncommon gastrointestinal (GI) symptoms, shedding a higher understanding of gut–lung axis affecting the progression of COVID-19 \citep{li_microbiome_2021}.

@@

In addition to negative controls, already a common practice in any sequencing assay and in particular in metataxonomics (see \secref{sssec:_intro_metataxonomics}), positive controls can be a way to circumvent the lack of golden standards, either through the spike of the samples with a known amount of a specific DNA/RNA or though the sequencing of samples with known composition and abundance. Well-characterised reference standards and controls are needed to ensure shotgun metagenomics assay quality and stability over time \citep{chiu_clinical_2019, mcintyre_comprehensive_2017}. Most available metagenomic reference materials are highly tailored to a specific application. For example, the ZymoBIOMICS Microbial Community Standard\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}} is the first commercially available standard for microbiomics and metagenomics studies, providing mock a mock community with defined composition and abundance consisting of Gram-positive, Gram-negative and yeast. It is useful to determine the limit of detection of an assay, and the effectiveness and biases of a given protocol. Standards with a more limited spectrum of organisms are also available, such as the National Institute of Standards and Technology (NIST)\footnote{\url{https://www.nist.gov/}} reference materials for mixed microbial DNA detection, which contain only bacteria. Thus, these materials may not apply to untargeted shotgun metagenomics analyses.

@@

First generation sequencing technology (see \secref{ssec:_intro_1st_gen_seq}) produces far fewer reads than second generation sequencing technology (see \secref{ssec:_intro_2nd_gen_seq}, but individual reads are longer (500–1000 \ac{bp}). 
Assembly of Sanger data usually uses overlap-layout consensus (OLC) approaches, in which:

@@

These type of assemblers were very popular in the early 2010s, with assemblers such as Celera\footnote{\url{https://www.cbcb.umd.edu/software/celera-assembler}},  Genovo\footnote{\url{https://cs.stanford.edu/genovo }}, xGenovo\footnote{\url{http://xgenovo.dna.bio.keio.ac.jp/}} and BBAP\footnote{\url{http://homepage.ntu.edu.tw/~youylin/BBAP.html}} having been widely used \citep{myers_whole-genome_2000, hutchison_genovo_2010, afiahayati_extended_2013, lin_novo_2017}.
With the emergence of third-generation sequencing (see \secref{ssec:_intro_3rd_gen_seq}), OLC assemblers have been increasingly developed and adopted by the community to assembly long-read data. 
In the latest years, ra\footnote{\url{https://github.com/lbcb-sci/ra}}, raven\footnote{\url{https://github.com/lbcb-sci/raven}} and canu\footnote{\url{https://github.com/marbl/canu}}, the latter being a a fork of the Celera Assembler, have become staples in the community, showing good reliability and amassing over 3000 citations \citep{vaser_yet_2019, koren_canu_2017, wick_benchmarking_2021}.  %missing raven

@@

Metataxonomics (see \secref{sssec:_intro_metataxonomics}) is the most widely used technique for microbial diversity analysis \citep{hilton_metataxonomic_2016}, and due to its particularities, the analysis of this data is also very particular. Data analyses are mostly carried out through specialised pipelines that wrap and combine several tools, offering the possibility to follow a simple protocol with default configurations or choose between a plethora of different configurations to adjust for any particular needs. Quantitative Insights Into Microbial Ecology 2 (QUIIME2)\footnote{\url{https://qiime2.org/}} \citep{bolyen_reproducible_2019} has become the \textit{de facto} tool for metataxonomic analysis as a framework with an ever-growing suite of plugins and intuitive data visualisation tools for the assessment of results. Mothur \citep{schloss_introducing_2009} and UPARSE \citep{edgar_uparse_2013} are also a popular alternative although resulting outputs differing significantly between pipelines despite using the same inputs having been reported by \cite{marizzoni_comparison_2020}, with a magnitude that is comparable to differences in upstream sample treatment and sequencing procedures.  A typical workflow starts with quality filtering, error correction and removal of chimeric sequences. These quality control steps are followed by either taxonomic assignment of reads or a clustering step where reads are gathered into OTUs given their sequence identity, followed by statistical analysis to assess differences between given groups. Taxonomic assignment methods classify query sequences based on the best hit found in reference databases of annotated sequences, being heavily dependent on the completeness of the reference databases (see \secref{sssec:_intro_metataxonomics}). Classification is further limited by lack of species annotation in most reference databases \citep{westcott_novo_2015}. Alternatively, the same approach of direct taxonomic classification, without OTU clustering, can be followed as with genomic and shotgun metagenomic data, given that the databases include \ac{rRNA} sequences.

>>>>

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. 
\ac{PCR} and \ac{RT-PCR} are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of and demand for prompt information about viral loads, viral sequence data and potential antiviral resistance information \citep{cassedy_virus_2021}. 
Syndromic testing (see \secref{sssec:_intro_bacterial}) is now fully integrated into standard testing practises of many clinical laboratories \citep{dien_bard_panels_2020}. 
The limitations of these assays include the absence of detection of off-target pathogens, a lack of full susceptibility information, cost, and false positive results. 
\ac{qPCR} remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 5.png}
\caption{\textbf{The three revolutions in sequencing technology that have transformed the landscape of bacterial genome sequencing.} The first-generation, also known as Sanger sequencers, is represented by the ABI Capillary Sequencer (Applied Biosystems). During the sequencing reaction, at each nucleotide incorporation event, a fluorescently labelled \ac{ddNTP} is incorporated, terminating the elongation of the DNA molecule. The resulting electropherogram for sequencing reaction is below, and is read from left to right. The second-generation, also known as high-throughput sequencers, is represented by MiSeq, a 4-channel sequencer, and NextSeq, a 2-channel sequencer (Illumina), both sequencing by synthesis instruments. For both instruments, the loaded flowcell is sequenced in massive parallel reactions, with each nucleotide incorporation emitting a light signal that is captured and latter basecalled into a fastq file, with indication of the confidence of the call, presented below. In a 4-channel instrument each nucleotide has its own marker (A: yellow, T: green, C: red, G: blue) but in a 2-channel instrument only 2 markers exist (A: green plus red, T: green, C: red, G: no marker). These instruments allow the sequencing of both ends of the DNA fragment. Lastly, the third-generation, also known as long-read sequencers, is represented by the Pacific Bioscience BS sequencer and Oxford Nanopore MinION sequencer. In the first, immobilised polymerases in a \ac{SMRT} Cell incorporating nucleotides with identifying fluorescent labels. In the latter, a nanopore embedded in a solid-state membrane causes a change in an ionic current across the membrane each time a nucleotide is pushed through the pore. This difference in potential is then used for basecalling. Adapted from \cite{hagemann_overview_2015, loman_twenty_2015,goodwin_coming_2016, wang_nanopore_2021, metzker_sequencing_2010, xu_recent_2020}}
\label{fig:figure5}
\end{figure*}

@@

The basis of Pacific Biosciences sequencers is known as \ac{SMRT}, which takes place in single-use \ac{SMRT} Cells. 
These contain multiple immobilised polymerases, which, after binding to an adaptor sequence, begin replication incorporating nucleotides with identifying fluorescent labels. 
The sequence of fluorescence pulses is recorded into a movie which is then converted into a nucleotide sequence. 
After the polymerase completes replication of one DNA strand, it continues to sequence the opposite adapter and second strand. 
As a result, multiple passes of the same template can be generated depending on the lifetime of the polymerase \citep{hoang_long-reads-based_2022, loman_twenty_2015}. 
This technology has accuracy comparable with the Illumina systems but requires a higher initial investment cost, are much larger machines in comparison with the benchtop counterparts, and have much lower throughput and longer library preparation protocols \citep{hoang_long-reads-based_2022, wenger_accurate_2019}.

@@

\ac{WGS} is becoming one of the most widely used applications of microbial genome sequencing. 
The major advantage of \ac{WGS} is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:_intro_metagenomics}). 
In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:_intro_bacterial}) or viral (see \secref{sssec:_intro_viral}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures (see Section \ref{sec:_intro_bioinformatics}), but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, nature_reviews_genetics_genomic_2019, goodwin_coming_2016}.

@@

\ac{WGS} has been used in the routine laboratory workflow when typing of pathogens by a method having the highest possible discriminatory power is required either through \ac{SNP} or \ac{cg/wg MLST} analysis, for example during hospital outbreaks \citep{tagini_bacterial_2017}.

The implementation of \ac{WGS} in routine diagnostics requires several adaptations in the laboratory workflow, from the ‘wet’ laboratory part (extraction, library preparation, sequencing), to the ‘dry’ bioinformatics part where genomic data is analysed and its results interpreted by specialised personnel \citep{rossen_practical_2018}.

@@

Most notably, \ac{WGS} has become a common tool in surveillance and infection prevention, allowing for pathogen identification and tracking, establishing transmission routes and outbreak control \citep{lo_genomics_2020}. 
In bacterial infections, initiatives such as Pathogenwatch\footnote{\url{https://pathogen.watch/}} offers a web-based platform for \ac{AMR} analysis and phylogeny generation of \textit{Campylobacter}, \textit{Klebsiella}, \textit{Neisseria gonorrhoeae}, \textit{Staphylococcus aureus}, and \textit{Salmonella Typhi} \citep{afolayan_overcoming_2021}. 
The Center for Genomic Epidemiology website\footnote{\url{https://www.genomicepidemiology.org/}} offers services for phylogenetic tree building and \ac{AMR} prediction. 
Chewie Nomenclature Server\footnote{\url{https://chewbbaca.online/}} allows users to share genome-based gene-by-gene typing schemas and to maintain a common nomenclature, simplifying the comparison of results \citep{mamede_chewie_2021}. 
Enterobase\footnote{\url{https://enterobase.warwick.ac.uk/}} allows for the analysis and visualisation of genomic variation within enteric bacteria \citep{zhou_enterobase_2020}. 
Microreact\footnote{\url{https://microreact.org/}}, from the same developers as Pathogenwatch, combines clustering, geographical and temporal data into an interactive visualisation with trees, maps, timelines and tables for a multitude of microorganisms, both bacterial and viral \citep{argimon_microreact_nodate}. 
Particularly for viruses, GISAID\footnote{\url{https://www.gisaid.org/}}  promotes the rapid sharing of data from all influenza viruses and the coronavirus causing COVID-19, including the genetic sequences and related clinical and epidemiological data \citep{shu_gisaid_2017}. 
ViPR\footnote{\url{https://www.viprbrc.org/}} provides access to sequence records, gene and protein annotations, immune epitopes, 3D structures, host factor data, and other data types for over 14 viral families, including \textit{Coronaviridae}, from which \ac{SARS-CoV-2} belongs to, and \textit{Faviviridae}, the family of Dengue and Zika virus \citep{pickett_virus_2012}. 
INSaFLU\footnote{\url{https://insaflu.insa.pt/}} supplies public health laboratories and influenza researchers with a web-based suite for effective and timely influenza and \ac{SARS-CoV-2} laboratory surveillance, identifying the type and subtype/lineage, detection of putative mixed infections and intra-host minor variants \citep{borges_insaflu_2018}.

@@

Molecular barcoding approaches can be combined with second-generation high-throughput sequencing to achieve unprecedented depths of coverage in microbial community profiling, being defined as metataxonomics. 
For profiling bacterial species, the most popular approach is 16S \ac{rRNA} gene sequencing, an ~1500 base pair gene coding for a catalytic RNA that is part of the 30S ribosomal subunit. 
Traditionally, the variable regions of the 16S \ac{rRNA} gene (V-regions) are targeted, or ranges thereof (V1-V2, V1-V3, V3-V4, V4, V4-V5, V6-V8, and V7-V9), and are specific to bacterial genus (96\%) and for some, even species (87.5\%), \citep{srinivasan_use_2015, abellan-schneyder_primer_2021}. 
Moreover, dedicated 16S databases that include near full-length sequences for a large number of strains and their taxonomic placements exist, such as RDP\footnote{\url{http://rdp.cme.msu.edu/}}, Greengenes\footnote{\url{https://greengenes.secondgenome.com/}}, silva\footnote{\url{https://www.arb-silva.de/}} and NCBI's 16S ribosomal RNA project\footnote{\url{https://www.ncbi.nlm.nih.gov/refseq/targetedloci/}} \citep{cole_ribosomal_2009, desantis_greengenes_2006, pruesse_silva_2007}. 
The sequence from an unknown strain can be compared against the sequences in these databases, after very closely related sequences are grouped into \ac{OTUs},  an operational definition used to classify groups of closely related individuals. This allows for deducing of likely taxonomy, with the assumption that sequences of $>$95\% identity represent the same genus, whereas sequences of $>$97\% identity represent the same species \citep{schloss_introducing_2005}. 
In addition, NCBI also provides the 23S ribosomal RNA project for Bacteria and Archaea metataxonomics.

Furthermore, it must necessarily account for intragenomic variation between 16S gene copies. 
Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. 
Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 \ac{bp}) gene \citep{johnson_evaluation_2019}. 
The emergence of third generating sequencing technologies (see \secref{ssec:_intro_3rd_gen_seq}) allows for this limitation to be overcome but currently, only a fraction of the databases includes complete 16S \ac{rRNA} sequences.

@@

\ac{SMg} can offer relatively unbiased pathogen detection and characterisation. The capacity to detect all potential pathogens — bacteria, viruses, fungi and parasites — in a sample has great potential utility in the diagnosis of infectious disease \citep{chiu_clinical_2019}, potentially able to provide genotyping, antimicrobial resistance and virulence profiling in a single methodological step. This comes with the cost of producing massive amounts of information that require expert handling and processing, as well as capable computational infrastructures \citep{couto_critical_2018, rossen_practical_2018}.

Clinical applications of \ac{SMg} derive its roots from the use of microarrays (see \secref{ssec:_intro_current_standards}), where it was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of second-generation sequencing technology and its high throughput of genomic data at a competitive price that made the sequencing of all genomic content, DNA and/or RNA) if a clinical sample a viable possibility for diagnostics (see \secref{ssec:_intro_2nd_gen_seq}) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of \ac{SMg} was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since many other reports of successful application of \ac{SMg} in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

In public health microbiology, \ac{SMg} combined with transmission network analysis allowed the investigation and quick action on the food supply of the 2013 outbreak of \ac{STEC} strain O104:H4 from faecal specimens obtained from patients \citep{loman_culture-independent_2013}. A similar approach was followed in the detection of \textit{Salmonella enterica} subsp. \textit{enterica} serovar Heidelberg from faecal samples in two though to be unrelated outbreaks in the United States of America, as well as the \textit{in situ} abundance and level of intrapopulation diversity of the pathogen, and the possibility of co-infections with \textit{Staphylococcus aureus}, overgrowth of commensal \textit{Escherichia coli}, and significant shifts in the gut microbiome during infection relative to reference healthy samples \citep{huang_metagenomics_2017}. More recently, shotgun metagenomic sequencing has evidenced alterations in the gut microbiota of a subset of COVID-19 patients that present the uncommon gastrointestinal (GI) symptoms, shedding a higher understanding of gut–lung axis affecting the progression of COVID-19 \citep{li_microbiome_2021}.

@@

In addition to negative controls, already a common practice in any sequencing assay and in particular in metataxonomics (see \secref{sssec:_intro_metataxonomics}), positive controls can be a way to circumvent the lack of golden standards, either through the spike of the samples with a known amount of a specific DNA/RNA or though the sequencing of samples with known composition and abundance. Well-characterised reference standards and controls are needed to ensure the quality and stability of the \ac{SMg} assay over time \citep{chiu_clinical_2019, mcintyre_comprehensive_2017}. Most available metagenomic reference materials are highly tailored to a specific application. For example, the ZymoBIOMICS Microbial Community Standard\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}} is the first commercially available standard for microbiomics and metagenomics studies, providing mock a mock community with defined composition and abundance consisting of Gram-positive, Gram-negative and yeast. It is useful to determine the limit of detection of an assay, and the effectiveness and biases of a given protocol. Standards with a more limited spectrum of organisms are also available, such as the \ac{NIST}\footnote{\url{https://www.nist.gov/}} reference materials for mixed microbial DNA detection, which contain only bacteria. Thus, these materials may not apply to untargeted \ac{SMg} analyses.

@@

First generation sequencing technology (see \secref{ssec:_intro_1st_gen_seq}) produces far fewer reads than second generation sequencing technology (see \secref{ssec:_intro_2nd_gen_seq}, but individual reads are longer (500–1000 \ac{bp}). 
Assembly of Sanger data usually uses \ac{OLC} approaches, in which:

@@

These type of assemblers were very popular in the early 2010s, with assemblers such as Celera\footnote{\url{https://www.cbcb.umd.edu/software/celera-assembler}},  Genovo\footnote{\url{https://cs.stanford.edu/genovo }}, xGenovo\footnote{\url{http://xgenovo.dna.bio.keio.ac.jp/}} and BBAP\footnote{\url{http://homepage.ntu.edu.tw/~youylin/BBAP.html}} having been widely used \citep{myers_whole-genome_2000, hutchison_genovo_2010, afiahayati_extended_2013, lin_novo_2017}.
With the emergence of third-generation sequencing (see \secref{ssec:_intro_3rd_gen_seq}), \ac{OLC} assemblers have been increasingly developed and adopted by the community to assembly long-read data. 
In the latest years, ra\footnote{\url{https://github.com/lbcb-sci/ra}}, raven\footnote{\url{https://github.com/lbcb-sci/raven}} and canu\footnote{\url{https://github.com/marbl/canu}}, the latter being a a fork of the Celera Assembler, have become staples in the community, showing good reliability and amassing over 3000 citations \citep{vaser_yet_2019, koren_canu_2017, wick_benchmarking_2021}.

@@

Metataxonomics (see \secref{sssec:_intro_metataxonomics}) is the most widely used technique for microbial diversity analysis \citep{hilton_metataxonomic_2016}, and due to its particularities, the analysis of this data is also very particular. Data analyses are mostly carried out through specialised pipelines that wrap and combine several tools, offering the possibility to follow a simple protocol with default configurations or choose between a plethora of different configurations to adjust for any particular needs. Quantitative Insights Into Microbial Ecology 2 (QUIIME2)\footnote{\url{https://qiime2.org/}} \citep{bolyen_reproducible_2019} has become the \textit{de facto} tool for metataxonomic analysis as a framework with an ever-growing suite of plugins and intuitive data visualisation tools for the assessment of results. Mothur \citep{schloss_introducing_2009} and UPARSE \citep{edgar_uparse_2013} are also a popular alternative although resulting outputs differing significantly between pipelines despite using the same inputs having been reported by \cite{marizzoni_comparison_2020}, with a magnitude that is comparable to differences in upstream sample treatment and sequencing procedures.  A typical workflow starts with quality filtering, error correction and removal of chimeric sequences. These quality control steps are followed by either taxonomic assignment of reads or a clustering step where reads are gathered into \ac{OTUs} given their sequence identity, followed by statistical analysis to assess differences between given groups. Taxonomic assignment methods classify query sequences based on the best hit found in reference databases of annotated sequences, being heavily dependent on the completeness of the reference databases (see \secref{sssec:_intro_metataxonomics}). Classification is further limited by lack of species annotation in most reference databases \citep{westcott_novo_2015}. Alternatively, the same approach of direct taxonomic classification, without \ac{OTUs} clustering, can be followed as with genomic and shotgun metagenomic data, given that the databases include \ac{rRNA} sequences.

\ac{OTUs} clustering methods can be categorised into: (1) computationally expensive hierarchical methods that cluster sequences based on a distance matrix measuring the difference between each pair of sequences, (2) less expensive heuristic methods cluster sequences into \ac{OTUs} based on a pre-defined threshold, generally, with a sequence being selected as a seed and the rest of the sequences being analysed sequentially and added to existing or new clusters according to the defined threshold, and (3) model based clustering methods that do not rely on a pre-defined and fixed threshold, defining \ac{OTUs} based on a soft threshold and carrying out the clustering process based on methods such as an unsupervised probabilistic Bayesian clustering algorithm \citep{hao_clustering_2011}. These methods offer the possibility to cluster sequences based on criteria that do not depend on reference databases and are especially useful in less characterised microbial communities or with a high representation of uncultured microbes. Due to the assumptions made with this strategy, it is sensitive to under or overestimation of the number of \ac{OTUs} in a sample as defining a threshold to accurately cluster sequences is difficult \citep{westcott_novo_2015}.
